Global Warming Distributed Data Analysis

### 2016-03-28 report/article.tex

\documentclass{vldb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{siunitx}

@@

\subsection{Data Sources}
We have access to more than one hundred years of weather data provided by the NOAA (National Oceanic and Atmospheric Administration) organization. The raw data is publicly accessible\footnote{\href{ftp://ftp.ncdc.noaa.gov/pub/data/noaa/}{Noaa public data}}, so anyone interested in analyzing the climate of the past years can use them. The data is organized by year and grouped by ISD (Integrated Surface Data) stations. They also provide a \texttt{perl} script that helps in reading and viewing the data. NOAA also publishes monthly reports on the climate status (like temperature, precipitation, sea level) and compares the data obtained for the last month/year to the averages calculated in the past, accompanied by a software suite to repeat or modify the analysis at home. While this tools and datasets are helpful, there are currently no publicly available tools to perform custom data analysis on the whole raw dataset available. In particular, considering the size of the uncompressed data (\~205 GB) it may be necessary to use a cluster to make complex computations that require more time. \\

@@

@@

\subsection{Data Analysis Environment}
The technological environment we decided to use for this project consist of:
\begin{itemize}
    \item \textbf{HDFS}\footnote{\href{Hadoop Distributed File System}{http://hadoop.apache.org/}}: a distributed file system that allows the developers to store and retrieve large quantity of data between multiple machines as if it was a single one;
    \item \textbf{Spark}\footnote{\href{Apache Spark}{http://spark.apache.org/}}: programming framework to develop applications for large scale data processing on horizontally scalable clusters.
\end{itemize}

@@

The data stored in the distributed file system is grouped in folders by year. As shown in figure \ref{fig:dataSize} the size of the data to analyze is very different between the years. Inside each folder, there is a file for each station that reported measurements. HDFS is not very efficient when working with many small files\cite{zhang2012improving}, as in our case. The solution we opted to use is to merge the small files in bigger files that have the same size as Hadoop's default block size. To merge the files we used filecrusher\footnote{\href{Hadoop Filecrusher}{https://github.com/edwardcapriolo/filecrush}}. This tool merges small files, located inside an HDFS instance, into bigger files with the size of a multiplier of an HDFS block. We have run this tool one time per each folder of the data. The number of total files was reduced by an order of magnitude of $10^2$.\\

@@

\textbf{SURFsara Hadoop cluster}
All the computation are executed on the SURFsara Hadoop cluster \footnote{\href{SURFsara Hadoop Cluster}{https://userinfo.surfsara.nl/systems/hadoop}} . The cluster is one of the largest Hadoop clusters for scientific research in the Netherlands. It is composed of 170 data/compute nodes with a total 1370 CPU-cores for parallel processing. The system has also a distributed file system that has a capacity of 2.3 PB. The cluster is shared and used by different users that needs an HPC infrastructure for their research.

@@

\subsection{Regional five year period analysis}
The aim of this analysis is to calculate the temperature statistics (mean, maximum, minimum and standard deviation) on every area of the earth for which we have data. After that we can compare the situation for the same areas in different time frames and show it with a world map view (as shown in \ref{sec:map}). \\
The first problem is that while the regions we want to compare are the same size and uniformly distributed around the earth, the stations are not. Furthermore a lot of stations are added and removed during the flow of the years, this means that the data of a new station couldn't be compared with any data of the previous years even if in the previous years there was a station very close to the new added station. So we defined regions of size 1\degree latitude and 1\degree longitude. This is approximately the area covered by the region of Holland in the Netherlands. Then we assigned each measurement (for each one there is the coordinates reported) to a region. \\
Our analysis proceeded by loading five consecutive years of measurements at a time, group them in small regions and, for each region, calculate the arithmetic mean, the minimum, the maximum temperature and the standard deviation. This was done for all years from 1901 to 2015 (the year 2016 was excluded from the analysis due to incomplete data).\\
\subsection{Median}
Another value we wanted to extract from the datasets is the median. We implemented the code to calculate it, but we didn't have enough time to test it and obtain the results for all the datasets. The problem with calculating the median is that it is not a commutative and associative operation (the median of medians is not the median of the entire group). This means that there is no easy way to use the reduce function to combine results computed in different machines in a distributed way. The simple way to calculate the median is to sort the desired data and look at the value in the middle. The problem with this solution is that it's not scalable. Indeed, applied to datasets in the order of GB in sizes this operation is very slow. Work has been done by various researchers to tackle this kind of problems and improve the efficiency of non associative operations (\cite{liu2009computing}).

@@

\subsection{Temperature yearly trends analysis}
In this type of analysis we focused on some particular large areas of on the earth. We selected the regions that for which we have more and better distributed data available. The selected regions are:
\begin{itemize}
    \item United States: from coordinate N\ang{50;;} W\ang{128;;} to N\ang{24;;} W\ang{65;;};
    \item Europe: from coordinate N\ang{71;;} W\ang{11;;} to N\ang{35;;} E\ang{41;;};
    \item Australia: from coordinate S\ang{9;;} E\ang{111;;} to S\ang{44;;} E\ang{156;;};
\end{itemize}
For each region we calculated minimum, maximum, average and standard deviation of every year until 2015. The year range is chosen based on the availability of data.

@@

\subsection{Interactive map}
\label{sec:map}
The first tool we worked on is a dynamic map that allows to visualize the difference between temperature values (average, maximum, minimum and standard deviation) between two datasets. Each dataset is created from the analysis one described in section \ref{sec:exp} and are composed of the aggregated results of the five yearly analysis in which we divided the surface in small regions of 1\degree latitude and 1\degree longitude. \\
We used the map provided by Google\footnote{\href{Google Maps}{https://developers.google.com/maps/}}. We created a custom heatmap that shows on each region of the map different colors to show an increase or decrease for a particular value in that region (shades of blue for reduction and red for a gain). The interface can be seen in figure \ref{fig:interface}

@@

\subsection{Zonal Analysis}
In this section we report the results of the regional analysis done for: Europe, United States and Australia. We start considering the amount of data available from each region for each year.

Now we move on to consider the average analysis on the three selected regions.

@@

\newpage
\appendix
\section{Stations distribution} \label{App:AppendixA}
Here is an overview of the stations for some selected years. We decided to show only years in which the number and position of the stations is significantly different than in the previous or following years.

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1907}
\caption{Wheater stations used in 1907. In the first years we only have measurements coming from stations in Finland}
\label{fig:stations1907}
\end{figure}
\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1930}
\caption{Wheater stations used in 1930. Here we can see the stations cover the most part of central Europe. The problem we encountered is that there is no continuity in the stations that are used between consecutive years, so it's difficult to make comparisons.}
\label{fig:stations1930}
\end{figure}
\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1945}
\caption{Wheater stations used in 1945. Around this year we start to have a relevant number of stations that cover most part of the surface.}
\label{fig:stations1945}
\end{figure}
\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1968}
\caption{Wheater stations used in 1968. In the late 60s we notice a sudden drop in the earth coverage compared to the previous years. We took this into account when making the analysis.}
\label{fig:stations1968}
\end{figure}
\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1972}
\caption{Wheater stations used in 1972. As shown in figure \ref{fig:stations}, we noticed a sudden drop in the number of stations for this year. In particular we observe the absence of stations from east Europe and the ex Soviet Union Area.}
\label{fig:stations1972}
\end{figure}
\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1980}
\caption{Wheater stations used in 1980. From this year onwards we have a large number of stations covering most part of the surface.}
\label{fig:stations1980}
\end{figure}
\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations2014}
\caption{Wheater stations used in 2014.}
\label{fig:stations2014}
\end{figure}

\clearpage
%\end{document}  % This is where a 'short' article might terminate

% ensure same length columns on last page (might need two sub-sequent latex runs)
\balance

@@

% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{ieeetr}
\bibliography{vldb_sample}  % vldb_sample.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references

%Generated by bibtex from your ~.bib file.  Run latex,
%then bibtex, then latex twice (to resolve references).

>>>>

\documentclass{vldb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{siunitx}
\usepackage{placeins}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
 \usepackage[backend=bibtex]{biblatex}

\bibliography{vldb_sample}  % vldb_sample.bib is the name of the Bibliography in this case

@@

\subsection{Data Sources}
We have access to more than one hundred years of weather data provided by the NOAA (National Oceanic and Atmospheric Administration) organization. The raw data is publicly accessible\cite{noaa}, so anyone interested in analyzing the climate of the past years can use them. The data is organized by year and grouped by ISD (Integrated Surface Data) stations. They also provide a \texttt{perl} script that helps in reading and viewing the data. NOAA also publishes monthly reports on the climate status (like temperature, precipitation, sea level) and compares the data obtained for the last month/year to the averages calculated in the past, accompanied by a software suite to repeat or modify the analysis at home. While this tools and datasets are helpful, there are currently no publicly available tools to perform custom data analysis on the whole raw dataset available. In particular, considering the size of the uncompressed data (\~205 GB) it may be necessary to use a cluster to make complex computations that require more time. \\

@@

\FloatBarrier

@@

\subsection{Data Analysis Environment}
The technological environment we decided to use for this project consist of:
\begin{itemize}
    \item \textbf{HDFS}\cite{hadoop}: a distributed file system that allows the developers to store and retrieve large quantity of data between multiple machines as if it was a single one;
    \item \textbf{Spark}\cite{spark}: programming framework to develop applications for large scale data processing on horizontally scalable clusters.
\end{itemize}

@@

The data stored in the distributed file system is grouped in folders by year. As shown in figure \ref{fig:dataSize} the size of the data to analyze is very different between the years. Inside each folder, there is a file for each station that reported measurements. HDFS is not very efficient when working with many small files\cite{zhang2012improving}, as in our case. The solution we opted to use is to merge the small files in bigger files that have the same size as Hadoop's default block size. To merge the files we used filecrush\cite{filecrush}. This tool merges small files, located inside an HDFS instance, into bigger files with the size of a multiplier of an HDFS block. We have run this tool one time per each folder of the data. The number of total files was reduced by an order of magnitude of $10^2$.\\

@@

\textbf{SURFsara Hadoop cluster}
All the computation are executed on the SURFsara Hadoop cluster \cite{surfsara}. The cluster is one of the largest Hadoop clusters for scientific research in the Netherlands. It is composed of 170 data/compute nodes with a total 1370 CPU-cores for parallel processing. The system has also a distributed file system that has a capacity of 2.3 PB. The cluster is shared and used by different users that needs an HPC infrastructure for their research.

\FloatBarrier

@@

\subsection{Regional five year period analysis}
The aim of this analysis is to calculate the temperature statistics (mean, maximum, minimum and standard deviation) on every area of the earth for which we have data. After that we can compare the situation for the same areas in different time frames and show it with a world map view (as shown in \ref{sec:map}). \\
The first problem is that while the regions we want to compare are the same size and uniformly distributed around the earth, the stations are not. Furthermore a lot of stations are added and removed during the flow of the years, this means that the data of a new station couldn't be compared with any data of the previous years even if in the previous years there was a station very close to the new added station. So we defined regions of size 1\degree latitude and 1\degree longitude. This is approximately the area covered by the region of Holland in the Netherlands. Then we assigned each measurement (for each one there is the coordinates reported) to a region. \\
Our analysis proceeded by loading five consecutive years of measurements at a time, group them in small regions and, for each region, calculate the arithmetic mean, the minimum, the maximum temperature and the standard deviation. This was done for all years from 1901 to 2015 (the year 2016 was excluded from the analysis due to incomplete data).\\
\subsubsection{Median}
Another value we wanted to extract from the datasets is the median. We implemented the code to calculate it, but we didn't have enough time to test it and obtain the results for all the datasets. The problem with calculating the median is that it is not a commutative and associative operation (the median of medians is not the median of the entire group). This means that there is no easy way to use the reduce function to combine results computed in different machines in a distributed way. The simple way to calculate the median is to sort the desired data and look at the value in the middle. The problem with this solution is that it's not scalable. Indeed, applied to datasets in the order of GB in sizes this operation is very slow. Work has been done by various researchers to tackle this kind of problems and improve the efficiency of non associative operations (\cite{liu2009computing}).

\subsection{Zonal Analysis}
In this type of analysis we focused on some particular large areas of on the earth. We selected the regions that for which we have more and better distributed data available. The selected regions are:
\begin{itemize}
    \item United States: from coordinate N\ang{50;;} W\ang{128;;} to N\ang{24;;} W\ang{65;;};
    \item Europe: from coordinate N\ang{71;;} W\ang{11;;} to N\ang{35;;} E\ang{41;;};
    \item Australia: from coordinate S\ang{9;;} E\ang{111;;} to S\ang{44;;} E\ang{156;;};
\end{itemize}
For each region we calculated minimum, maximum, average and standard deviation of every year until 2015. The year range is chosen based on the availability of data.

@@

\FloatBarrier

@@

\subsection{Interactive map}
\label{sec:map}
The first tool we worked on is a dynamic map that allows to visualize the difference between temperature values (average, maximum, minimum and standard deviation) between two datasets. Each dataset is created from the analysis one described in section \ref{sec:exp} and are composed of the aggregated results of the five yearly analysis in which we divided the surface in small regions of 1\degree latitude and 1\degree longitude. \\
We used the map provided by Google\cite{maps}. We created a custom heatmap that shows on each region of the map different colors to show an increase or decrease for a particular value in that region (shades of blue for reduction and red for a gain). The interface can be seen in figure \ref{fig:interface}

@@

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{increaseGlobal}
\caption{Map that shows the increase of the minimum temperature globally between 1976 and 2015.}
\label{fig:interface}
\end{figure}

\subsection{Zonal Analysis}
In this section we report the results of the regional analysis done for: Europe, United States and Australia.

@@

\clearpage

\appendix
\section{Stations distribution} \label{App:AppendixA}
Here is an overview of the stations for some selected years. We decided to show only years in which the number and position of the stations is significantly different than in the previous or following years.

\begin{figure}[tbh]

\includegraphics[width=1\linewidth]{stations1907}
\caption{Wheater stations used in 1907. In the first years we only have measurements coming from stations in Finland}
\label{fig:stations1907}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1930}
\caption{Wheater stations used in 1930. Here we can see the stations cover the most part of central Europe. The problem we encountered is that there is no continuity in the stations that are used between consecutive years, so it's difficult to make comparisons.}
\label{fig:stations1930}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1945}
\caption{Wheater stations used in 1945. Around this year we start to have a relevant number of stations that cover most part of the surface.}
\label{fig:stations1945}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1968}
\caption{Wheater stations used in 1968. In the late 60s we notice a sudden drop in the earth coverage compared to the previous years. We took this into account when making the analysis.}
\label{fig:stations1968}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1972}
\caption{Wheater stations used in 1972. As shown in figure \ref{fig:stations}, we noticed a sudden drop in the number of stations for this year. In particular we observe the absence of stations from east Europe and the ex Soviet Union Area.}
\label{fig:stations1972}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1980}
\caption{Wheater stations used in 1980. From this year onwards we have a large number of stations covering most part of the surface.}
\label{fig:stations1980}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations2014}
\caption{Wheater stations used in 2014.}
\label{fig:stations2014}
\end{figure}

\FloatBarrier
\clearpage
%\end{document}  % This is where a 'short' article might terminate

% ensure same length columns on last page (might need two sub-sequent latex runs)

@@

% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references

%Generated by bibtex from your ~.bib file.  Run latex,
%then bibtex, then latex twice (to resolve references).
\balance
\printbibliography
